Finally, functional class scoring enrichment analysis of Infinium DNA methylation data

### 2023-11-03 figs/GSE158422_simulate_summary.Rmd
results

```{r,summary,fig.width=12,fig.height=3}

par(mfrow=c(1,3))

@@

@@

barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, main="20 genes per set")
grid()

legend("topleft", inset=.1, c("Precision","Recall","F1"), fill=cols, cex=0.8)

@@

barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, main="50 genes per set")
grid()

@@

barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, main="100 genes per set")
grid()

>>>>

```{r,summary,fig.width=5,fig.height=4}

@@

x %>% kbl(caption="sim 20 res") %>% kable_paper("hover", full_width = F)

@@

barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="20 genes per set")
grid()

legend("topleft", inset=.1, c("Precision","Recall","F1"), fill=cols, cex=0.7)

@@

x %>% kbl(caption="sim 50 res") %>% kable_paper("hover", full_width = F)

barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="50 genes per set")
grid()

@@

x %>% kbl(caption="sim 100 res") %>% kable_paper("hover", full_width = F)

barplot(t(x),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="100 genes per set")
grid()

### 2023-11-03 manuscript/manuscript.Rmd
results

To assess these methods, we adopted a simulated data approach.
EPIC Infinium array data derived from non-cancerous lung tissue of 37 cancer patients was used for our
evaluations.
We randomly sampled datasets to serve as control and cases (no replacement), with group sizes varying between
3 and 12.
One thousand random gene sets were created with sizes varying between 20 and 100 members.
Randow gene sets were used to avoid problems caused by the large overlap among real gene sets.
Throughout our evaluations, 50 gene sets were selected to be differentially methylated, 25 with increased and 25
with decreased methylation.
From these gene sets, half of the members were selected.
For those selected genes, half of the annotated probes were selected.
We adjusted the m-values in the case group by a specified amount, which we call the "delta", which we varied
between 0.1 and 0.4.
Following the incorporation of methylation changes to selected probes in the case group, the data underwent
enrichment analysis using `GSAmeth` approach, or one of the seven FCS methods outlined above.
Enrichment results were then filtered for FDR<0.05 and correct direction of methylation change.
Differentially methylated gene sets observed were compared to gene sets selected to be the ground truth, so that
the number of true positives (TP), false positives (FP) and false negatives (FN) could be determined.
At each setting for group size, delta and gene set size, 20 replications were conducted each with a set seed
to ensure reproducibility.
Mean TP, FP and FN values were determined, and then precision, recall and F1 scores were calculated.

@@

Simulations were conducted using the seven FCS methods (LAT, LTT, LAW, LAM, ALT, ALW, ALM) and GSAmeth,
with group sizes of 3-12 and delta values between 0.1 and 0.5, and random gene sets with sizes 20, 50
and 100.
Precision and recall were calculated for these tests (Figure 1).
As expected, precision and recall values were better when group sizes and delta values were larger.

>>>>

To assess these methods, we adopted a simulated data approach.
EPIC Infinium array data derived from non-cancerous lung tissue of 37 cancer patients was used for our
evaluations.
We randomly sampled datasets to serve as control and cases (no replacement), with group sizes varying between
3 and 12.
One thousand random gene sets were created with sizes varying between 20 and 100 members.
Random gene sets were used to avoid problems caused by the large overlap among real gene sets.
Throughout our evaluations, 50 gene sets were selected to be differentially methylated, 25 with increased and 25
with decreased methylation.
From these gene sets, half of the members were selected.
For those selected genes, half of the annotated probes were selected.
We adjusted the m-values in the case group by a specified amount, which we call the "delta", which we varied
between 0.1 and 0.4.
Following the incorporation of methylation changes to selected probes in the case group, the data underwent
enrichment analysis using `GSAmeth` approach, or one of the seven FCS methods outlined above.
Enrichment results were then filtered for FDR<0.05 and correct direction of methylation change.
Differentially methylated gene sets observed were compared to gene sets selected to be the ground truth, so that
the number of true positives (TP), false positives (FP) and false negatives (FN) could be determined.
At each setting for group size, delta and gene set size, 20 replications were conducted each with a set seed
to ensure reproducibility.
Mean TP, FP and FN values were determined, and then precision, recall and F1 scores were calculated.

@@

Simulations were conducted using the seven FCS methods (LAT, LTT, LAW, LAM, ALT, ALW, ALM) and GSAmeth,
with group sizes of 3-12 and delta values between 0.1 and 0.5, and random gene sets with sizes 20, 50
and 100.
True positives, false positives and false negatives were used to calculate overall precision and recall
at these three gene set sizes (Figure 1).
We found that overall recall was positively associated with the size of the gene sets considered.
At 20 genes per set, GSA had the highest overall performance, with F1 of 0.25, followed by LAW with 0.06.
At 50 genes per set, GSA and LAM achieved similar F1 values of ~0.39.
At 100 genes per set, LAM achieved the highest F1 performance of 0.70 followed by LAT with 0.61, while
GSA scored only 0.42.
This result suggests that FCS methods perform better than ORA with gene set sizes >50.
Performance results for GSA and LAM methods with 100 genes per set shows that recall was higher for most
simulations, particularly with lower group sizes.
Taken together, this result suggests that LAM offers improved sensitivity over existing approaches when
the gene sets are larger than 50 members.

 in Figure 2.
This 
As expected, precision and recall values were better when group sizes and delta values were larger.
